---
title: "Comparing normalisations"
output: html_document
---
Load required packages:
```{r}
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(knitr)
library("clusterProfiler", lib = "~/Rlibs")
```
Read in raw counts data/control-normalised data/median-normalised data and proceed..(this r markdown has been randomly put together today so it doesn't go through them individually in a nice neat way)
```{r}
rawcounts = read.delim("all_nonorm.count.txt")
controlnorm = read.delim("all_controlnorm.count_normalized.txt", stringsAsFactors = F)
mediannorm = read.delim("all.count_normalized.txt", stringsAsFactors = F)
# Convert count columns to numeric - do for all of them
numeric_cols <- c("T0", "D7", "PSU", "PSL")
rawcounts[numeric_cols] <- lapply(rawcounts[numeric_cols], function(x) as.numeric(as.character(x)))
```
pre-processing: add filter for total guide dropout, add pseduocount and log-transform data
```{r}
totaldropout <- rawcounts %>%
  filter(T0 == 0 & D7 == 0 & PSU == 0 & PSL == 0)
filtered_rawcounts <- anti_join(rawcounts, totaldropout, by = "sgRNA")
countsMat <- filtered_rawcounts[, c("T0", "D7", "PSU", "PSL")]
logCounts = log2(countsMat + 0.5)
#plot mean counts and SD per guide - Miks way
mm = rowMeans(logCounts)
ss = apply(logCounts, 1, sd)
plot(mm, ss, pch='.', xlab = "Mean counts per guide", ylab="Standard deviation per guide")
#same thing again but my way
stats <- logCounts %>%
  mutate(mean_log = rowMeans(.),
         sd_log = apply(., 1, sd))
#Combine with sgRNA identifiers for reference
stats$sgRNA <- filtered_rawcounts$sgRNA
stats$Gene <- filtered_rawcounts$Gene
```
plot mean v SD and then identify non-targeting controls on the plot
```{r}
ggplot(stats, aes(x = mean_log, y = sd_log)) +
  geom_point(alpha = 0.4) +
  theme_minimal() +
  labs(
    title = "Mean–SD Relationship of log2 Counts",
    x = "Mean log2(count + 0.5)",
    y = "Standard deviation (log2 scale)")
#plot again with controls identified
ggplot(stats_filtered, aes(x = mean_log, y = sd_log, color = ControlType)) +
  geom_point(alpha = 0.4, size = 1) +
  geom_smooth(method = "loess", se = FALSE, color = "black", linewidth = 1) +
  scale_color_manual(values = c("Targeting sgRNA" = "grey50",
                                "Non-targeting control" = "red")) +
  theme_minimal(base_size = 14) +
  labs(
    title = "Mean–SD Relationship of log2 Counts with Negative Controls Highlighted",
    x = "Mean log2(count + 0.5)",
    y = "Standard deviation (log2 scale)",
    color = "Guide type"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "top",
    panel.grid.minor = element_blank())
```

Compare dispersion before and after normalisation for each

```{r}
cat("Pre-normalization SD range:", range(ss, na.rm = TRUE), "\n")
cat("Post-normalization SD range:", range(ss1, na.rm = TRUE), "\n")
cat("Pre-normalization mean SD:", mean(ss, na.rm = TRUE), "\n")
cat("Post-normalization mean SD:", mean(ss1, na.rm = TRUE), "\n")
cat("Variance of SDs pre-normalization:", var(ss, na.rm = TRUE), "\n")
cat("Variance of SDs post-normalization:", var(ss1, na.rm = TRUE), "\n")
```
