---
title: "CRISPR Screen Normalisation QC"
author: "Caroline Stokes"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# Overview

This report walks through the main count tables used in the analysis and visualises the effect of each normalisation strategy.

The workflow covers:

1. Raw non-normalised counts with a strict `<10 reads -> 0` filter
2. Control-normalised counts
3. Median-normalised counts
4. Median-normalised counts with the same strict `<10 reads -> 0` filter
5. A direct SD comparison between raw and median-normalised data

# Load packages

```{r libraries}
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(VennDiagram)
```

# Helper functions

```{r helpers}
coerce_numeric_cols <- function(df, numeric_cols) {
  df[numeric_cols] <- lapply(df[numeric_cols], function(x) as.numeric(as.character(x)))
  df
}

replace_zeros_with_pseudocount <- function(df, numeric_cols, pseudocount = 1) {
  df %>%
    mutate(across(all_of(numeric_cols), ~ ifelse(. == 0, pseudocount, .)))
}

apply_strict_filter <- function(df, numeric_cols, threshold = 10) {
  df %>%
    mutate(across(all_of(numeric_cols), ~ ifelse(. < threshold, 0, .)))
}

remove_total_dropouts <- function(df, numeric_cols, sgrna_col = "sgRNA") {
  totaldropout <- df %>%
    filter(if_all(all_of(numeric_cols), ~ . == 0))
  
  filtered <- anti_join(df, totaldropout, by = sgrna_col)
  
  list(
    filtered = filtered,
    totaldropout = totaldropout
  )
}

make_log_counts <- function(df, numeric_cols) {
  counts_mat <- as.matrix(df[, numeric_cols])
  log2(counts_mat)
}

plot_density_base <- function(logCounts, main_title = "Sample density plots", xlim_range = c(-3, 15)) {
  plot(density(logCounts[, 1]), lwd = 2, xlim = xlim_range, main = main_title)
  for (i in 1:ncol(logCounts)) {
    lines(density(logCounts[, i]), col = i, lwd = 2)
  }
  legend("topleft", colnames(logCounts), fill = 1:ncol(logCounts), bty = "n")
}

plot_qc_panels <- function(logCounts, sample_title = "") {
  mm <- rowMeans(logCounts)
  ss <- apply(logCounts, 1, sd)
  
  par(mfrow = c(2, 2))
  plot(mm, ss, pch = ".", xlab = "Mean counts per guide", ylab = "Standard deviation per guide",
       main = paste("Mean-SD plot", sample_title))
  
  plot_density_base(logCounts, main_title = paste("Sample density plots", sample_title))
  boxplot(logCounts, xlab = "", main = paste("Sample boxplots", sample_title))
  barplot(colSums(logCounts > 0), main = paste("Non-zero sgRNAs per sample", sample_title))
  par(mfrow = c(1, 1))
  
  invisible(list(mean_per_guide = mm, sd_per_guide = ss))
}

summarise_counts <- function(df, numeric_cols, label = "dataset") {
  tibble(
    dataset = label,
    n_guides = nrow(df),
    total_reads = paste(colSums(df[, numeric_cols]), collapse = ", ")
  )
}
```

# 1. Raw non-normalised counts (strict filtering)

```{r raw-load}
setwd("/home/stoca738/Hugheslab_project/Caroline_Mageck/H1/")
rm(list = ls()[!ls() %in% c(
  "coerce_numeric_cols", "replace_zeros_with_pseudocount", "apply_strict_filter",
  "remove_total_dropouts", "make_log_counts", "plot_density_base",
  "plot_qc_panels", "summarise_counts"
)])

counts_raw <- read.delim("allnonorm.count.txt")
raw_cols <- c("T0", "D7", "UL", "LL")
counts_raw <- coerce_numeric_cols(counts_raw, raw_cols)
```

## Guides per gene in the raw library

```{r raw-guides-per-gene}
guides_per_gene <- counts_raw %>%
  group_by(Gene) %>%
  summarise(total_guides = n(), .groups = "drop")

head(guides_per_gene)
```

## Apply strict filtering and remove total dropouts

```{r raw-strict-filter}
counts_raw_strict <- apply_strict_filter(counts_raw, raw_cols, threshold = 10)
raw_dropout_result <- remove_total_dropouts(counts_raw_strict, raw_cols, sgrna_col = "sgRNA")

totaldropout_raw <- raw_dropout_result$totaldropout
counts_raw_filtered <- raw_dropout_result$filtered

nrow(totaldropout_raw)
nrow(counts_raw_filtered)
```

## Add pseudocount and inspect distributions

```{r raw-qc}
counts_raw_filtered_pc <- replace_zeros_with_pseudocount(counts_raw_filtered, raw_cols, pseudocount = 1)
logCounts_raw <- make_log_counts(counts_raw_filtered_pc, raw_cols)
raw_stats <- plot_qc_panels(logCounts_raw, sample_title = "- Raw non-normalised")
```

## Raw summary

```{r raw-summary}
summarise_counts(counts_raw_filtered_pc, raw_cols, label = "Raw strict-filtered")
```

# 2. Control-normalised counts

```{r control-load}
setwd("/home/stoca738/Hugheslab_project/Caroline_Mageck/H1/")
counts_ctrlnorm <- read.delim("all_counts_controlnorm/all_controlnorm.count")
ctrl_cols <- c("T0", "D7", "PSU", "PSL")
counts_ctrlnorm <- coerce_numeric_cols(counts_ctrlnorm, ctrl_cols)
```

## Add pseudocount and inspect distributions

```{r control-qc}
counts_ctrlnorm_pc <- replace_zeros_with_pseudocount(counts_ctrlnorm, ctrl_cols, pseudocount = 1)
logCounts_ctrl <- make_log_counts(counts_ctrlnorm_pc, ctrl_cols)
ctrl_stats <- plot_qc_panels(logCounts_ctrl, sample_title = "- Control-normalised")
```

## Control-normalised summary

```{r control-summary}
summarise_counts(counts_ctrlnorm_pc, ctrl_cols, label = "Control-normalised")
```

# 3. Median-normalised counts

```{r median-load}
setwd("/home/stoca738/Hugheslab_project/Caroline_Mageck/H1/count_all/")
counts_mednorm <- read.delim("all.count_normalized.txt")
med_cols <- c("T0", "D7", "PSU", "PSL")
counts_mednorm <- coerce_numeric_cols(counts_mednorm, med_cols)
```

## Add pseudocount and inspect distributions

```{r median-qc}
counts_mednorm_pc <- replace_zeros_with_pseudocount(counts_mednorm, med_cols, pseudocount = 1)
logCounts_med <- make_log_counts(counts_mednorm_pc, med_cols)
med_stats <- plot_qc_panels(logCounts_med, sample_title = "- Median-normalised")
```

## Median-normalised summary

```{r median-summary}
summarise_counts(counts_mednorm_pc, med_cols, label = "Median-normalised")
```

# 4. Median-normalised counts with strict filtering

```{r median-strict}
counts_mednorm_strict <- apply_strict_filter(counts_mednorm, med_cols, threshold = 10)
med_dropout_result <- remove_total_dropouts(counts_mednorm_strict, med_cols, sgrna_col = "sgRNA")

totaldropout_med <- med_dropout_result$totaldropout
counts_mednorm_filtered <- med_dropout_result$filtered

nrow(totaldropout_med)
nrow(counts_mednorm_filtered)
```

## Add pseudocount and inspect distributions

```{r median-strict-qc}
counts_mednorm_filtered_pc <- replace_zeros_with_pseudocount(counts_mednorm_filtered, med_cols, pseudocount = 1)
logCounts_med_strict <- make_log_counts(counts_mednorm_filtered_pc, med_cols)
med_strict_stats <- plot_qc_panels(logCounts_med_strict, sample_title = "- Median-normalised strict-filtered")
```

## Median-normalised strict summary

```{r median-strict-summary}
summarise_counts(counts_mednorm_filtered_pc, med_cols, label = "Median-normalised strict-filtered")
```

# 5. Direct SD comparison: raw vs median-normalised

This section compares the per-guide standard deviation before and after median normalisation using the strict-filtered datasets.

```{r sd-comparison}
ss_raw <- raw_stats$sd_per_guide
ss_med <- med_strict_stats$sd_per_guide

cat("Pre-normalization SD range:", range(ss_raw, na.rm = TRUE), "\n")
cat("Post-normalization SD range:", range(ss_med, na.rm = TRUE), "\n")
cat("Pre-normalization mean SD:", mean(ss_raw, na.rm = TRUE), "\n")
cat("Post-normalization mean SD:", mean(ss_med, na.rm = TRUE), "\n")
cat("Variance of SDs pre-normalization:", var(ss_raw, na.rm = TRUE), "\n")
cat("Variance of SDs post-normalization:", var(ss_med, na.rm = TRUE), "\n")
```

## Visual comparison of SD distributions

```{r sd-density-comparison}
plot(density(ss_raw, na.rm = TRUE), lwd = 2,
     main = "Per-guide SD distribution: raw vs median-normalised",
     xlab = "Per-guide SD")
lines(density(ss_med, na.rm = TRUE), col = 2, lwd = 2)
legend("topright",
       legend = c("Raw strict-filtered", "Median-normalised strict-filtered"),
       col = c(1, 2), lwd = 2, bty = "n")
```

# 6. Optional combined summary table

```{r combined-summary}
bind_rows(
  summarise_counts(counts_raw_filtered_pc, raw_cols, label = "Raw strict-filtered"),
  summarise_counts(counts_ctrlnorm_pc, ctrl_cols, label = "Control-normalised"),
  summarise_counts(counts_mednorm_pc, med_cols, label = "Median-normalised"),
  summarise_counts(counts_mednorm_filtered_pc, med_cols, label = "Median-normalised strict-filtered")
)
```

# Notes

- The raw count table uses `UL` and `LL` sample names.
- The normalised count tables use `PSU` and `PSL` sample names.
- Zero values are replaced with a pseudocount of 1 before log2 transformation.
- The strict filter sets counts `<10` to zero before dropout removal.
